Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs

### 2018-10-22 wads/paper.tex
added folder for wads version, copied eurocg paper into llncs format

% for section-numbered lemmas etc., use "numberwithinsect"
\documentclass[a4paper,english,numberwithinsect]{eurocg18}

% the recommended bibstyle
\bibliographystyle{plainurl}

%------------------------------------------------------------------- 
%if unwanted, comment out or use option "draft"
\usepackage{todonotes}
\usepackage{microtype}
\usepackage[noend, linesnumbered]{algorithm2e}
\usepackage{amsmath}
\usepackage{color}
\usepackage{cite}
\usepackage{pifont}
\usepackage{enumerate}
\usepackage[basic]{complexity}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{graphicx}
%\usepackage{cutwin}

% Line numbers are helpful for refereeing
\usepackage[mathlines]{lineno}
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
\renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}
%\linenumbers

%helpful if your graphic files are in another directory
%\graphicspath{{./graphics/}}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\newcommand\polylog{{\rm polylog}}

% Author metadata::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs}
%{Maximal Symmetric Partial Edge Drawings With Degree $ > 2$}
%optional, in case that the title is too long; 
%the running title should fit into the top page column
%\titlerunning{MaxSPEDs with degree $ > 2$}

%% Please provide for each author the \author and \affil macro, 
%even when authors have the same affiliation, i.e. for each 
%author there needs to be the  \author and \affil macros
\author[1]{Michael H\"oller}
\author[1]{Fabian Klute}
\author[1]{Soeren Nickel}
\author[1]{Martin~N\"ollenburg}
\author[1]{Birgit Schreiber}

\affil[1]{Algorithms and Complexity Group, TU Wien, Vienna, Austria\\ \texttt{[fklute|noellenburg]@ac.tuwien.ac.at, first.lastname@student.tuwien.ac.at}}
%soeren.nickel@tuwien.ac.at}}
%\affil[2]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{fklute@ac.tuwien.ac.at}}
%\affil[3]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%\affil[4]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{noellenburg@ac.tuwien.ac.at}}
%\affil[5]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%mandatory. First: Use abbreviated first/middle names. 
%Second (only in severe cases): Use first author plus 'et. al.'
\authorrunning{M. Höller, F. Klute, S. Nickel, M. N\"ollenburg, and B. Schreiber}

\ArticleNo{50}

% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{document}

\maketitle

@@

\begin{wrapfigure}[11]{r}{0.45\textwidth}
	\centering
	\includegraphics{export_full}
	\hfill
	\includegraphics{export_alpha_nontouching}
	\caption{A straight-line graph drawing (a) and a maximum-ink symmetric partial edge drawing (b) of the same graph.}
	\label{fig:examples}
\end{wrapfigure}

The idea of drawing edges only partially has been formalized in graph drawing as follows~\cite{bk-ecbe-12}. 
A \emph{partial edge drawing (PED)} is a graph drawing that maps vertices to points and edges to pairs of crossing-free edge stubs of positive length pointing towards each other.
These edge stubs are obtained by erasing one contiguous central piece of the straight-line segment connecting the two endpoints of each edge.
In other words each straight-line edge is divided into three parts, of which only the two outer ones are drawn (see Fig.~\ref{fig:examples}).
More restricted and better readable~\cite{blmt-pedhmitc-16} variations of PEDs are \emph{symmetric} PEDs, in which both stubs of an edge must have the same length (see Fig.~\ref{fig:examples}(b)), and \emph{homogeneous} PEDs, in which the ratio of the stub length to the total edge length is the same for all edges.
Symmetric stubs facilitate finding adjacent vertices due to the identical stub lengths at both vertices, and symmetric homogeneous stubs additionally indicate the distance at which to find a neighboring vertex.
The natural optimization problem in this formal setting is \emph{ink maximization}, i.e., maximizing the total stub length, so that as much information as possible is given in the drawing while all crossings disappear in the negative background space.

@@

\section{Preliminaries}\label{sec:preliminaries}
%\begin{itemize}
%	\item basic definitions and notation
%	\item segment intersection graph
%	\item alle planaren Graphen können auftreten (insbes. Bäume und Kakteen)
%\end{itemize}
%Input: $k$-plane graph drawing $\Gamma$ of a graph $G$ with edge set $S={s_1, \dots, s_m}$ (set of segments)
%Intersection graph $C=(V,E)$ with $V=S$ and edges if two segments intersect 
%Consider maximum degree of $C$ -- here mostly 3.

Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight-line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. We use the terms edge in $ S $ and segment in $ \Gamma $ interchangeably. %, since our input is always a graph $G$ with a given drawing $ \Gamma $. 
Hence we can also interpret $S$ as a set of line segments.

@@

%\fabian{This part has to be adapted once the intro is fixed}
%
%We define a \emph{partial edge drawing} (\ped) $ \Gamma' $ of $ G $ as a drawing using the same embedding as $ \Gamma $, but edges are drawn as \emph{stubs}, i.e., we do not depict an edge $ s_i \in S $ by a line segment connecting the two vertices, but remove the middle part of the edges such that no two stubs intersect in $ \Gamma' $. We call a \ped $ \Gamma' $ \emph{symmetric} if for every edge $ s_i s_j \in S $ its two stubs have equal lengths in $ \Gamma' $. Finally we define the problem \maxsped as finding a symmetric \ped $ \Gamma' $ of $ G $ where the sum of the lengths of the stubs is maximized. This is also known as ink maximization~\cite{bcgkmn-pped-17}.
%
%\martin{Maybe we don't need the above paragraph as PED, SPED, maxSPED are all somewhat defined in the introduction}

@@

\begin{theorem}\label{thm:hard}
	\maxsped is \NP-hard even for 3-plane graph drawings.
\end{theorem}

%\renewcommand\windowpagestuff{%
%	\includegraphics{Reduction_visual}
%	\captionof{figure}{Example of three variable gadgets and\\ a satisfied clause gadget. Dotted parts do not belong\\ to the SPED.}
%	\label{fig:gadget_connection}
%}
%\opencutleft
\begin{proof}
%		\begin{cutout}{2}{0pt}{0.44\textwidth}{15}
	We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as in Bruckdorfer's sketch of the hardness proof for general \maxsped~\cite{b-sgh-15}. 
%	Bruckdorfer reduces from the more restricted problem \textsc{Planar Positive 1-in-3 Sat}.
Here we specify precisely the maximum ink contributions of all gadgets needed for a satisfying variable assignment. 
Our variable gadgets are cycles of edge pairs that admit exactly two maximum-ink states.
%Further we changed the variable gadgets to cycles of edge pairs, partly to guarantee, that the gadgets have the same ink value regardless of their corresponding truth values and partly for aesthetic reasons. 
Finally we construct clause gadgets consisting of three pairwise intersecting edges so that all crossings are between two edges only, while Bruckdorfer's gadgets have multiple edges intersecting in the same point.
	Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
	We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$, which has a vertex for each variable $x_i$ and a vertex for each clause $c_j$.
	Each clause vertex is connected to the three variables appearing in the clause.
	In the drawing of $H_\phi$ all variable vertices are placed on a horizontal line and the clause vertices connect to the adjacent variable vertices either from above or from below the horizontal line.
	In our reduction (see Fig.~\ref{fig:gadget_connection}) we mimic the drawing of $H_\phi$ by creating a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma_\phi$ has a SPED with ink at least $L$ if and only if $\phi$ is satisfiable.
%	Figure~\ref{fig:gadget_connection} shows a sketch of the gadgets.
%		\end{cutout}

		\begin{figure}
%	\begin{wrapfigure}[15]{r}{0.55\textwidth}

			\centering
\includegraphics{Reduction_visual}
\caption{Example of three variable gadgets and a satisfied clause gadget. Dotted parts do not belong to the SPED.}
\label{fig:gadget_connection}
%	\end{wrapfigure}		
		\end{figure}
	All segments in the gadgets are of length 5.
	We use pairs of intersecting segments, alternatingly colored red and green.
	The intersection point of each red-green segment pair is at distance 1 from an endpoint.
	Thus, the maximum amount of ink contributed by such a pair is 7 (one full segment of length 5 and the other one with two stubs of length 1 each).

	Each variable gadget is a cycle of segment pairs, with (at least) one pair for each occurrence of the variable in $\phi$, see Fig.~\ref{fig:gadget_connection}. 
	Observe that this cycle has exactly two ink-maximal SPEDs: either all red edges are full segments and all green edges are length-1 stubs or vice versa.
	We associate the configuration with green stubs and full red segments with the value \emph{true} and the configuration with full green segments and red stubs with the value \emph{false}.

@@

	Conversely, assume that we have a SPED with ink value $L$. 
	By construction, every red-green segment pair and every clause gadget must contribute its respective maximum ink value.
	In particular, each variable gadget is either in state \emph{true} or \emph{false}.
	By design of the gadgets it is straight-forward to verify that the corresponding truth assignment satisfies~$\phi$.
%
	% We claim that this truth assignment satisfies $\phi$.
	% For each clause, one segment must have full length and as argued before the path connecting that segment to a variable gadget must have all segments at even positions as length-1 stubs and all segments at odd positions as full segments.
	% In particular, the first segment of the literal path is fully drawn and therefore, in a SPED with ink $L$, the intersecting segment of the variable cannot be fully drawn.
	% In fact, it must be drawn as length-1 stubs for otherwise the variable gadget would not contribute its maximum ink value.
	% But then by construction, the selected variable value sets our literal to \emph{true}.
\end{proof}

%\newpage %delete later
\section{Two polynomial special cases}

%\begin{itemize}
%	\item dynamic programming
%	\item table entries $T_i(s)$ for $i = 1, \dots, \deg(s)$
%	\item $i$ corresponds to increasing stub lengths
%	\item stub lengths $l_i(s)$ for $i = 1, \dots, \deg(s)$ (not the pair of stubs)
%\end{itemize}

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. 
Now we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms. Let $ C = (V,E) $ be the intersection graph of a given drawing $\Gamma$ of a graph $ G $ as defined in Section~\ref{sec:preliminaries}. 
Let $ u \in V $ and $ \delta  = \deg(u) $. 
Then for the corresponding segment $ s(u) \in S $ there are $ \delta + 1 $ relevant stub pairs including the whole segment, see Fig.~\ref{fig:tree}. 
Let $ \ell_1(u), \dots, \ell_\delta(u) \in \mathbb{R}_+$ be the stub lengths induced by the intersection points of $s(u)$ with the segments of the neighbors of $u$,  sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole segment $ s(u) $.

\begin{figure}[tbp]
	\centering
 	\includegraphics{tree_illustration}
	\caption{A segment $s(u)$ with five intersecting segments and the different induced stub lengths. The boxed stub lengths are considered in $\solshort(u)$ and do not affect $p(u)$.} % The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
	\label{fig:tree}
\end{figure}

%Computing the intersection graph requires to identify all the intersections in $ \Gamma $. Using a standard sweep line algorithm \cite{}\fabian{add citation} this requires $ O(|S|\log |S| + I) $ where $ I $ is the number of crossings. Building the graph then takes $ O(|S|) $ time. When computing the intersection we can compute the stub lengths and their order as well.
%
%\martin{time to build graph should be $O(S+I)$; but maybe this preparation work should go to preliminaries so that we can assume the intersection graph as input}
%\fabian{add to prelims}

@@

% \begin{figure}[tbp]
% 	\centering
%  	\includegraphics{tree_illustration}
% 	\caption{One edge $ s(u) $ with intersecting segments. The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
% 	\label{fig:tree}
% \end{figure}

Here we assume that $ C =(V,E) $ is a rooted tree of maximum degree $ k $. 
We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. 
For each vertex $u \in V$ we compute and store the maximum ink values $T_i(u)$ for $i = 0, \dots, \delta$ with $\delta = \deg(u)$ for the subtree rooted at $u$ such that $s(u)$ is drawn as a pair of stubs of length $\ell_i(u)$.
For $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. 
For $ u \in V $  %with parent $ p(u) \in V $ 
let $ i_p $ be the index of the stub length $\ell_{i_p}(u)$ induced by the intersection point of $s(u)$ and $s(p(u))$.
%the shorter length between one of $ s(u) $'s endpoints and the intersection point between $ s(u) $ and $ s(p(u)) $. 
We define the following two values, which allow us to categorize the stub lengths into those not affecting the stubs of the parent and those that do affect the parent:
%Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ \delta $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{\delta}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:tree} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
%\[	
	\solshort(u) = \max\{T_1(u),\dots,T_{i_p}(u)\}\\
	\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}.
%\]
\end{align*}
Figure~\ref{fig:tree} highlights the stub lengths that are considered in $\solshort(u)$. We recursively define % for a vertex $ u $ and stub length $ \ell_i(u) $:
\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i(u) \text{ intersects } s(v) \\
		\sollong(v) & \text{otherwise.}
	\end{cases}
\end{align}
The correctness of Recurrence~(\ref{rec:tree}) follows by induction. 
For a leaf $u$ in $C$ the set $c(u)$ is empty and the correctness of $T_i(u)$ is immediate.
Further, $\solshort(u) = T_1(u)$ and $\sollong(u) = T_0(u)$ are set correctly for the parent $p(u)$.
%Let $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. 
For an inner vertex $ u $ with degree $\delta$ we can assume by the induction hypothesis that the values $ \solshort(v) $ and $ \sollong(v) $ are computed correctly for all children $ v \in c(u) $. 
Each value $ T_i(u) $ for $ 0 \leq i \leq \delta $ is then the stub length $ \ell_i(u) $ plus the sum of the maximum ink we can achieve among the children subject to the stubs of $u$ being drawn with length $\ell_i(u)$. 
Setting $ \sollong(u) $ and $ \solshort(u) $ as above yields the two maximum ink values that are relevant for $ p(u) $.
%\begin{lemma}
%	\label{lem:tree_correctness}	
%	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \sollong(r) = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
%\end{lemma}
%\begin{proof}
%	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.
%	and $ t = \sollong(r) $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
%\end{proof}

Recurrence~(\ref{rec:tree}) can be solved naively in $ O(mk^2) $ time, where $m = |V|$. 
Using the order on the stub lengths  we can improve this to $ O(mk) $ time by computing all $ T_i(u) $ for one $ u \in V $ in $O(k)$ time.
Let $u \in V$ be a vertex with degree $\deg(u) = \delta$. 
The values $T_0(u) = \ell_0(u) + \sum_{v\in c(u)} \solshort(v)$ and $T_1(u) = \ell_1(u) + \sum_{v\in c(u)} \sollong(v)$ for the whole segment $s(u)$ and the shortest stubs can be computed in $O(k)$ time each.
Now $T_{j+1}(u)$ can be computed from $T_j(u)$ in $O(1)$ time as follows.
Let $v_j $ be the neighbor of $u$ that induces stub length $\ell_j(u)$ and assume $v_j \ne p(u)$.
In $T_j(u)$ we could still count the value $\sollong(v_j)$, but in $T_{j+1}(u)$ the stub length of $u$ implies that $v_j$ can contribute only to $\solshort(v_j)$. 
Then $T_{j+1}(u) = T_j(u) - \sollong(v_j) + \solshort(v_j)$.
If $v_j=p(u)$, then the two values $T_j(u)$ and $T_{j+1}(u)$ are equal as the corresponding change in stub length has no effect on the children of $u$.
Computing $\solshort(u)$ and $\sollong(u)$ clearly takes $O(k)$ time.

% \begin{lemma}
% 	\label{lem:compute_vertex}
% 	Let $ C = (V,E) $ be a tree with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every $ u \in V $ with degree $ \delta $. We can compute $ \solshort(u) $ and $ \sollong(u) $ in $ O(\delta) $ time and space, if the values $ \sollong(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
% \end{lemma}
% \begin{proof}
% 	 Let $ u \in V $ be a vertex with degree $ \delta < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children.
% %	 The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order.
% 	First we compute $ T_0(u) $ and $ T_1(u) $, this can be done since they represent the longest (whole edge) and shortest stub lengths.
% 	 Now assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $, hence we want to compute $ T_{j+1}(u) $ next. Doing so only requires to consider the next stub-length $ \ell_{j + 1}(u) $, since all other affected by drawing $ s(u) $ with $ \ell_{j+1}(u) $ where already considered. Hence we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.
%
% 	Finally we need to compute $ \solshort(u) $ and $ \sollong(u) $, but this are just two maximizations over at most $ \delta + 1 $ values at most. This can be done in $ O(\delta) $ by simply sweeping over the values.
% \end{proof}

@@

% The correctness follows as in the case of a tree in Section~\ref{sec:tree} if the values computed for the block vertices is correct. This is the case since we only extended the cycle by attaching some leafs, which can be dealt with after cutting the cycle open as in \cite{bcgkmn-pped-17}. For the runtime we can again do the same as in Section~\ref{sec:tree}, but this time we have to also check cycles of length $ n_c $ which leads to a runtime of $ O(n_c\delta) $ for a block vertex $ u \in \mathcal V $ with degree $ \delta $ and $ n_c = |\textit{cycle}(u)| $. In case $ u $ is an articulation vertex we can solve it in $ O(\delta) $. As before the optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ can be found by backtracking from the root. Additionally the intersection graph can be computed in $ O(m\log m) $ where $ m $ is the number of edges in the original graph $ G $ and we obtain the following theorem.

>>>>

\documentclass[a4paper, USenglish]{llncs}

\bibliographystyle{plainurl}

\usepackage{todonotes}
\usepackage{graphicx}
\usepackage[basic]{complexity}
\usepackage{wrapfig}
\usepackage{subfig}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{amsmath, amssymb}

\title{Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs}
\author{Fabian Klute \and Soeren Nickel \and Martin~N\"ollenburg}
\institute{Algorithms and Complexity Group, TU Wien, Vienna, Austria \email{[fklute|noellenburg]@ac.tuwien.ac.at, soeren.nickel@tuwien.ac.at}}

@@

\begin{document}
	\maketitle

@@

	\begin{figure}
		\centering
		\subfloat{
			\includegraphics{figures/export_full}
		}
		\subfloat {
			\includegraphics{figures/export_alpha_nontouching}
		}	
		\caption{A straight-line graph drawing (a) and a maximum-ink symmetric partial edge drawing (b) of the same graph.}
		\label{fig:examples}
	\end{figure}

	The idea of drawing edges only partially has been formalized in graph drawing as follows~\cite{bk-ecbe-12}. 
	A \emph{partial edge drawing (PED)} is a graph drawing that maps vertices to points and edges to pairs of crossing-free edge stubs of positive length pointing towards each other.
	These edge stubs are obtained by erasing one contiguous central piece of the straight-line segment connecting the two endpoints of each edge.
	In other words each straight-line edge is divided into three parts, of which only the two outer ones are drawn (see Fig.~\ref{fig:examples}).
	More restricted and better readable~\cite{blmt-pedhmitc-16} variations of PEDs are \emph{symmetric} PEDs, in which both stubs of an edge must have the same length (see Fig.~\ref{fig:examples}(b)), and \emph{homogeneous} PEDs, in which the ratio of the stub length to the total edge length is the same for all edges.
	The natural optimization problem in this formal setting is \emph{ink maximization}, i.e., maximizing the total stub length, so that as much information as possible is given in the drawing while all crossings disappear in the negative background space.

@@

	\section{Preliminaries}\label{sec:preliminaries}

	Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. We use the terms edge in $ S $ and segment in $ \Gamma $ interchangeably. %, since our input is always a graph $G$ with a given drawing $ \Gamma $. 
	Hence we can also interpret $S$ as a set of line segments.

@@

@@

	\begin{theorem}\label{thm:hard}
		\maxsped is \NP-hard even for 3-plane graph drawings.
	\end{theorem}
	\begin{proof}
		We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as in Bruckdorfer's general proof of the hardness of \maxsped~\cite{b-sgh-15}. 
		Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
		We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$, which has a vertex for each variable $x_i$ and a vertex for each clause $c_j$.
		Each clause vertex is connected to the three variables appearing in the clause.
		In the drawing of $H_\phi$ all variable vertices are placed on a horizontal line and the clause vertices connect to the adjacent variable vertices either from above or from below the horizontal line.
		In our gadget-based reduction we mimic the drawing of $H_\phi$ by creating a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma_\phi$ has a SPED with ink at least $L$ if and only if $\phi$ is satisfiable.
		Figure~\ref{fig:gadget_connection} shows a sketch of the gadgets.
		\begin{figure}
			\centering
			\includegraphics{figures/Reduction_visual}
			\caption{Example of three variable gadgets and a satisfied clause gadget. Dotted parts do not belong to the SPED.}
			\label{fig:gadget_connection}
		\end{figure}
		All segments in the gadgets are of length 5.
		We use pairs of intersecting segments, alternatingly colored red and green.
		The intersection point of each red-green segment pair is at distance 1 from an endpoint.
		Thus, the maximum amount of ink contributed by such a pair is 7 (one full segment of length 5 and the other one with two stubs of length 1 each).

		Each variable gadget is a cycle of segment pairs, with at least one pair for each occurrence of the variable in $\phi$, see Fig.~\ref{fig:gadget_connection}. 
		Observe that this cycle has exactly two ink-maximal SPEDs: either all red edges are full segments and all green edges are length-1 stubs or vice versa.
		We associate the configuration with green stubs and full red segments with the value \emph{true} and the configuration with full green segments and red stubs with the value \emph{false}.

@@

		Conversely, assume that we have a SPED with ink value $L$. 
		By construction, every red-green segment pair and every clause gadget must contribute its respective maximum ink value.
		In particular, each variable gadget is either in state \emph{true} or \emph{false}.
		We claim that this truth assignment satisfies $\phi$.
		For each clause, one segment must have full length and as argued before the path connecting that segment to a variable gadget must have all segments at even positions as length-1 stubs and all segments at odd positions as full segments. 
		In particular, the first segment of the literal path is fully drawn and therefore, in a SPED with ink $L$, the intersecting segment of the variable cannot be fully drawn.
		In fact, it must be drawn as length-1 stubs for otherwise the variable gadget would not contribute its maximum ink value.
		But then by construction, the selected variable value sets our literal to \emph{true}.	
	\end{proof}

	\section{Two polynomial special cases}

	Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. 
	Now we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms. Let $ C = (V,E) $ be the intersection graph of a given drawing $\Gamma$ of graph $ G $ as defined in Section~\ref{sec:preliminaries}. 
	Let $ u \in V $ and $ \delta  = \deg(u) $. 
	Then for the corresponding segment $ s(u) \in S $ there are $ \delta + 1 $ relevant stub pairs including the whole segment, see Fig.~\ref{fig:tree}. 
	Let $ \ell_1(u), \dots, \ell_\delta(u) \in \mathbb{R}_+$ be the stub lengths induced by the intersection points of $s(u)$ with the segments of the neighbors of $u$,  sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole segment $ s(u) $.

	\begin{figure}[tbp]
		\centering
		\includegraphics{figures/tree_illustration}
		\caption{A segment $s(u)$ with five intersecting segments and the different induced stub lengths. The boxed stub lengths are considered in $\solshort(u)$ and do not affect $p(u)$.} % The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
		\label{fig:tree}
	\end{figure}

@@

	Here we assume that $ C =(V,E) $ is a rooted tree of maximum degree $ k $. 
	We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. 
	For each vertex $u \in V$ we compute and store the maximum ink values $T_i(u)$ for $i = 0, \dots, \delta$ with $\delta = \deg(u)$ for the subtree rooted at $u$ such that $s(u)$ is drawn as a pair of stubs of length $\ell_i(u)$.
	For $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. 
	For $ u \in V $  %with parent $ p(u) \in V $ 
	let $ i_p $ be the index of the stub length $\ell_{i_p}(u)$ induced by the intersection point of $s(u)$ and $s(p(u))$.

	We define the following two values, which allow us to categorize the stub lengths into those not affecting the stubs of the parent and those that do affect the parent:
	\begin{align*}
		\solshort(u) = \max\{T_1(u),\dots,T_{i_p}(u)\}\\ % \qquad
		\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}.
	\end{align*}
	Figure~\ref{fig:tree} highlights the stub lengths that are considered in $\solshort(u)$. We recursively define 
	\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
	\solshort(v) & \text{if } s(u) \text{ with length } \ell_i(u) \text{ intersects } s(v) \\
	\sollong(v) & \text{otherwise}
	\end{cases}.
	\end{align}
	The correctness of Recurrence~(\ref{rec:tree}) follows by induction. 
	For a leaf $u$ in $C$ the set $c(u)$ is empty and the correctness of $T_i(u)$ is immediate.
	Further, $\solshort(u) = T_1(u)$ and $\sollong(u) = T_0(u)$ are set correctly for the parent $p(u)$.
	%Let $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. 
	For an inner vertex $ u $ with degree $\delta$ we can assume by the induction hypothesis that the values $ \solshort(v) $ and $ \sollong(v) $ are computed correctly for all children $ v \in c(u) $. 
	Each value $ T_i(u) $ for $ 0 \leq i \leq \delta $ is then the stub length $ \ell_i(u) $ plus the sum of the maximum ink we can achieve among the children subject to the stubs of $u$ being drawn with length $\ell_i(u)$. 
	Setting $ \sollong(u) $ and $ \solshort(u) $ as above yields the two maximum ink values that are relevant for $ p(u) $.

	Recurrence~(\ref{rec:tree}) can be solved naively in $ O(mk^2) $ time, where $m = |V|$. 
	Using the order on the stub lengths  we can improve this to $ O(mk) $ time by computing all $ T_i(u) $ for one $ u \in V $ in $O(k)$ time.
	Let $u \in V$ be a vertex with degree $\deg(u) = \delta$. 
	The values $T_0(u) = \ell_0(u) + \sum_{v\in c(u)} \solshort(v)$ and $T_1(u) = \ell_1(u) + \sum_{v\in c(u)} \sollong(v)$ for the whole segment $s(u)$ and the shortest stubs can be computed in $O(k)$ time each.
	Now $T_{j+1}(u)$ can be computed from $T_j(u)$ in $O(1)$ time as follows.
	Let $v_j $ be the neighbor of $u$ that induces stub length $\ell_j(u)$ and assume $v_j \ne p(u)$.
	In $T_j(u)$ we could still count the value $\sollong(v_j)$, but in $T_{j+1}(u)$ the stub length of $u$ implies that $v_j$ can contribute only $\solshort(v_j)$. 
	Then $T_{j+1}(u) = T_j(u) - \sollong(v_j) + \solshort(v_j)$.
	If $v_j=p(u)$ the two values $T_j(u)$ and $T_{j+1}(u)$ are equal as the corresponding change in stub length has no effect on the children of $u$.
	Computing $\solshort(u)$ and $\sollong(u)$ clearly takes $O(k)$ time.

@@

	\begin{theorem}
		Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a cactus with maximum degree $ k \in \mathbb{N} $, problem \maxsped can be solved in $ O(mk + m \log m) $ time and space.
	\end{theorem}

	\bibliography{paper}
\end{document}
